Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-14 paper.tex
Light resectioning

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

@@

% % % % % % % % % % % % % % % % % %
\subsection{SMC backwards-in-time}\label{par:description}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Likelihoods} \label{par:liks}
We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard''). 
As discussed in the Introduction, in many cases these events are not what is 
estimated by inference methods, which produce, more fundamentally, a collection
of inheritance relationships (edges) between ancestral haplotypes (nodes).
Our goal, therefore, is to decompose the overall likelihood into the
per-node and edge contributions to enable convenient and efficient 
calculation. 
% , as well as graceful degradation in the presence of incomplete information.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{Flexibility}
% likelihood surface
% The last paragraph aims to highlight the flexibility of this approach 
% and the impact of the various
% degrees of completeness of the ARG on the likelihood.
% Note that the figures in this paragraph only serve this exact purpose,
% we are not claiming to have designed a new parameter estimation approach.
In line with the observation that the presence of locally unary nodes 
(see section \ref{par:recording}) is crucial for the equivalence between
the graph or local trees view of an ARG, one can gradually simplify down
a simulated ARG by gradually removing the various types of locally unary nodes.
Figure \ref{fig:lik-surface} shows the impact of these simplification steps
on the likelihood for a single simulated ARG.
The log likelihood ratio is shown to improve comparability.

>>>>

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{SMC backwards-in-time}\label{par:description}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{ARG likelihood} \label{par:liks}
We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard''). 
As discussed in the Introduction, in many cases these events are not what is 
estimated by inference methods, which produce, more fundamentally, a collection
of inheritance relationships (edges) between ancestral haplotypes (nodes).
Our goal, therefore, is to decompose the overall likelihood into the
per-node and edge contributions to enable convenient and efficient 
calculation. 
% , as well as graceful degradation in the presence of incomplete information.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{Illustrative examples}
% likelihood surface
% The last paragraph aims to highlight the flexibility of this approach 
% and the impact of the various
% degrees of completeness of the ARG on the likelihood.
% Note that the figures in this paragraph only serve this exact purpose,
% we are not claiming to have designed a new parameter estimation approach.
[JK: Need a bit more narraitive here to feed back into the setup 
in the intro. Computing a likelihood on any ARG, graceful degradation
of performance with incomplete input..]

In line with the observation that the presence of locally unary nodes 
(see section \ref{par:recording}) is crucial for the equivalence between
the graph or local trees view of an ARG, one can gradually simplify down
a simulated ARG by gradually removing the various types of locally unary nodes.
Figure \ref{fig:lik-surface} shows the impact of these simplification steps
on the likelihood for a single simulated ARG.
The log likelihood ratio is shown to improve comparability.
